---
title: "MATH244 Final Project Presentation"
author: "Danish Club: Anders Hansen and Miles Jensen"
subtitle: "04/30/2026"
execute:
  echo: true
format:
  revealjs:
    theme: default 
    footnote: "Final Project"
editor: 
  markdown: 
    wrap: 72
engine: knitr
---

```{r load-packages}
#| warning: false
#| echo: false
#| message: false
library(tidyverse)
library(ggplot2)
library(dplyr)
library(reticulate)
library(tidyverse)
library(stringr)
library(ggcorrplot)
set.seed(67)
reticulate::use_miniconda("~/Library/r-miniconda-arm64/bin/python3.12")
```

# Project Introduction

## Project Overview and Motivation {.smaller}

#### Motivation
- Determine if song success is driven by tangible characteristics of the music or if it is based on "unmeasureables" (i.e. marketing, luck, timing, "catchiness")
  - Using a sample of hit and non-hit songs releases between 2010 and 2019
  - We define success as appearance on the Billboard Hot 100 

- Why the 2010s?
  - This period represents formative years for many of us (childhood/teens)
  - The music was really good (some songs are still charting today)
  - Successful song criteria seemed to change a lot over this period

#### Goal:
- Predict success of 2010s songs based on song characteristics
  - Create a model that is over 50% accurate in differentiating hit songs from non-hit songs

## Data Sources {.smaller}

```{r load-data}
#| message: false
#| warning: false
#| echo: false
hits_data <- read_tsv("data/hits_dataset.csv")
nonhits_data <- read_tsv("data/nonhits_dataset.csv")
artists_data <- read_tsv("data/artists.csv")
```

[MusicOSet](https://marianaossilva.github.io/DSW2019/)

* An open and enhanced dataset of musical elements (music, albums, and artists) suitable for music data mining
* Explanatory Variables:
  * Spotify audio features
  * Spotify artist genres
* Outcome Variable:
  * Billboard Hot 100 Data

The three datasets we used:

1. hits_dataset.csv: Spotify audio features for songs that **did** chart (size: 11,959 x 24)
2. nonhits_dataset.csv: Spotify audio features for songs that **did not** chart made by hit artists (size: 899,068 x 24)
3. artists.csv: main genre associated with each artist (size: 11,518 x 8)

## Data Cleaning: artists {.smaller}

`main_genre`

* Very specific/niche 
* Searched for keywords to sort them into broader genres instead ('Rap/Hip Hop', 'Pop', 'Indie', etc.) and then the nicher genres we put into 'Other'
* Genres with less than 100 observations in the combined dataset we reclassified into 'Other' (prevent overfitting)

```{r artists-genre-broaden}
#| echo: false
artists_data_cleaned <- artists_data %>%
  # drop artists with no genre listed
  filter(main_genre != "-") %>% 
  # only keep the artist_id and main_genre
  select(c(artist_id, main_genre)) 

# get the broad genre for the most common niche genres
artists_data_cleaned <- artists_data_cleaned %>%
  mutate(
    genre = case_when(
      str_detect(tolower(main_genre), "hip hop|rap|trap|drill|grime|g funk|new jack swing|crunk") ~ "Hip Hop/Rap",
      str_detect(tolower(main_genre), "pop|dance|electropop|bubblegum|boy band|freestyle|idol") ~ "Pop",
      str_detect(tolower(main_genre), "indie|alternative|shoegaze|lo-fi|emo|dreamo") ~ "Indie",
      str_detect(tolower(main_genre), "country|outlaw|hillbilly") ~ "Country",
      str_detect(tolower(main_genre), "rock|punk|metal|grunge|garage|british invasion|jam band|deathcore|permanent wave") ~ "Rock/Metal",
      str_detect(tolower(main_genre), "electronic|edm|techno|house|dubstep|dnb|disco|electro|big room|big beat") ~ "Electronic/Dance",
      str_detect(tolower(main_genre), "soul|r&b|neo soul|funk|neo mellow|motown") ~ "Soul/R&B/Funk",
      # str_detect(tolower(main_genre), "jazz|blues|bebop") ~ "Jazz/Blues", # less than 100 observations in combined dataset
      # str_detect(tolower(main_genre), "gospel|ccm|worship|christian") ~ "Gospel/Christian", # less than 100 observations in combined dataset
      # str_detect(tolower(main_genre), "latin|reggae|salsa|cumbia|sertane|banda") ~ "Latin/Reggae", # less than 100 observations in combined dataset
      # str_detect(tolower(main_genre), "folk|bluegrass|americana") ~ "Folk", # less than 100 observations in combined dataset
      str_detect(tolower(main_genre), "classical|baroque|orchestra|chamber") ~ "Classical",
      str_detect(tolower(main_genre), "adult standards|easy listening|lounge|piano cover|sleep|classify|coverchill") ~ "Easy Listening",
      str_detect(tolower(main_genre), "soundtrack|broadway|show tunes|musical|hollywood") ~ "Soundtrack/Theater",
      str_detect(tolower(main_genre), "comedy|comic|karaoke|a cappella|fake") ~ "Comedy/Entertainment",
      str_detect(tolower(main_genre), "musica para ninos") ~ "Children's Music",
      TRUE ~ "Other"
    )
  )

# ids to remove -> catalogs added to spotify later 
ids_remove = c('2vDV0T8sxx2ENnKXds75e5', '3tx8fyu3c4OBP5nejYtUOb', '5xLSa7l4IV1gsQfhAMvl0U', '5uwRBFpU9SQ9s2Us0nlXkk', '48nwxUvPJZkm8uPa7xMzmj', '4KDyYWR7IpxZ7xrdYbKrqY', '304senzcXGpWqcpssNpmXJ', '3Ri72CuuQSCLLkDRJgniFU', '1ktvF9yU9iWgY0aGlKfTS4', '0mopUs7SL5szuPOcOsS73A', '1PqdKx88nAgPolRy079lMl', '6J7biCazzYhU3gM9j1wfid', '1XIbUq7y7kiMUhmp5dywGx', '24qtJegdRiX2TPRvPN6rzk', '5gl2M2G5Dn5XTEW30iMMoD', '6I3M904Y9IwgDjrQ9pANiB', '0WxOgeRxUt0MwPrI7A5atQ', '04qIJRFjTmvW5I1DMyGE1R', '0MPtuQaV2GiRdLjAkPOaan', '52GxmJdAcByy1ZyPivpUns', '3VhvGNnWfzmlEyiuhzdgf5', '36hwOoNPgnsKnhoMBYpJrJ', '3xIkJAnwLy7NwZpPwvaSPc', '1cHFz6lrt7KAsBV8j2Ny1g', '3DEveX6KVzhPDqq2VRuN14', '45rpaf8BR69NIPqmeAYb7l', '2BLpGstUHxDc6vHfBEiaXm', '11XZpCBI7ExNNqmQoSgcJU', '6cAtGdArQhrU7tTBoeWY3P', '5BaHqGtf6UAZnHfqdPaTDA', '47fwSDHFIfnFlDVbYkSYkm', '4hKkEHkaqCsyxNxXEsszVH', '4hKkEHkaqCsyxNxXEsszVH', '2qbLM1AGCjpzC3PcOFMFU5', '0Lf3tvw0Rzwr38KDrShstn', '0h9smro0z3HqUbD94jotU8')

ids_remove2 = c('3RwQ26hR2tJtA8F9p2n7jG', '5jNGQ7VOU87x5402JjhTtd', '1mNnxxnPfHQDOkFjnZmdkc', '2XzHxDFTNywDd5ReMhgxkg', '7dXeJ8kAqTqtvNWQNV3sdU', '6lr9AruL30Dwl82fDODq9u', '7ohSq6OUvkH1hkslgXfGMi', '1L56apXminRPJqnqBN0vhe', '3LfO03nEZMdWNHG2tLpMa0', '5fbhwqYYh4YwUoEs582mq5', '25J2zw5Yzn51UowSvGjUyJ', '0BPupuaqc7AXufhUoyvMrJ', '0PDqlUGugF5Jt6DZyATvfz', '7qaC0L1S5rm7017VN33Zf7', '1OxJzMLmR9l5zPLap9OxuO', '51hRQLOySCpDBsBL6C03IZ', '5P53Exo01miUunItQyR2La', '0is49Iy1V6aUJYQgUYpr6x', '2aXiJJHJei5BmCykxI37y0', '6m30rs1IQqnWqV5nKMpU7U', '0efkWb9xKhT6H0E7kgSjFK', '7Kfrmups2Z3ncDQmNS5jRc', '7xGGqA85UIWX1GoTVM4itC', '6lUcc15t11bnaPr2qOuFs2', '0j4pxInmdtF4bYQj0VVpbz', '71cyZ6pH0KrM2MdtO8RGmy', '5HRSYURV6LFFRCjAtiK0py', '6avGAzQTmqu8mfSmZrN7gT', '742YR9ZgF0tCO5juBfvPAU', '3dkbV4qihUeMsqN4vBGg93', '6XItHNeKXecnFpkJHOn5JL', '7GaxyUddsPok8BuhxN6OUW', '5PmZd7YOaUgQwhOja44fmZ', '6vuD08WKtmp1yc7kQx1rBm', '1l51Sc9HBh8fLXUDEjFhxU', '0F4yfY9manxJrgsfwbXdZa')

ids_remove3 = c('2AV6XDIs32ofIJhkkDevjm', '5pq13QK3UcktjZaDw6dJEc', '27XI002zivG9xuzrZ9dIb1', '1ThoqLcyIYvZn7iWbj8fsj', '438JBZR1AR0l04AzcYW9gy', '0JCxGVxsISZzJHJPUOtceB', '3Bhgbs9kIwl8SuU0eIqN6b', '6PhJqhvF42CZ42NUmxIKLB', '0IecGJbdBeYSOVtSPRehh5', '0wbrRIw2po5DV4LeaHw55S', '3mC1KCuZZSOlN8Z0M56VsV', '1EialtL6J5R7O0exgdgQ1w', '7z1zaOeWiHpl4SD2N9ZBgk', '6YPRXu1dazGYcSZv4HJEH4', '0g3EsNvUCwgpsdkAiIC50W', '5Pek6HAr0BoHD9P1RuA7d5', '3IKV7o6WPphDB7cCWXaG3E', '4gXMKWbaoOBxvxAMz9Zbg4', '0ApoVi8QEh35twomSXh9TX', '0CXNdUUhv5Cw42dr0wWrjd', '4nts0oxMT67lVUoi5Kjxrb', '2uA0g2bLN2rIGZDkx671Ig', '67ZoIyKhetbA0x8LHzUXuE', '0701Axu6yvUIoctaKMbDIZ', '4NPzR70jYNxmCapRCwwBkr', '052NpRSiz4SQTsFnvwMP6t', '2wjWgyrpWbWmd077lMg04d', '7lffJlv0nRl0sIsHDmo0SB', '2SFOkFzbmtkUJ5A6LcYPWB', '4EBkFWxQ0EP5Fvya15d3Af', '3Inrg8cs8oc4q8oPES4a6S', '55izsTrkzLJ9KYKynnBmc0', '6LK7EEHFApe9FZbFRFyBDN', '5dORjaew9A94A25xvTMiTg', '1mg8nnhxHsBOe2lJHQvGdH', '0En4EEcDMJ5kaUCf1aZ9js')

ids_remove4 = c('3AOPsb5oGscweRef0jCI1p', '3EkURHRbSZWjatYLo36NMS', '75jNCko3SnEMI5gwGqrbb8', '2RwV62DIxFGEn8aWO3iaMM', '5TbXjzD8tYgMD5JU2g2F8q', '4dNs4gsyisVixgpkjyvD8D', '2jV8AcZ8W3TfiqInUGWwlL', '60df5JBRRPcnSpsIMxxwQm', '3BLqcKdqgBI3oJvVg7Odz8', '3J11x63maeBtbx8zXgzftQ', '4amyFeBFFIbbe6yq3ODoYC', '7erYZsfekXYGWvup1bnkUZ', '4r2UnA39uIXexqO1mABayb', '64wglyu5BoOzwMXYjtAbfS', '0Z4CzYz9ieK8q9XiVMPkW5', '4jZOiPysIzYxbDcDEOZmhu', '59fzcTUsoozikLmRY1SwHS', '3koiLjNrgRTNbOwViDipeA', '6ohvzFHYmLd05kyYhLui5K', '6Bfy6QzadCXS92y0T8dDZF', '1VPlW2qDDesSK0ZE0u7JXk', '1QAGLCom3FHTTiuRFsjzOj', '2iy394hZHUSLDMIxAnrz8x', '1WY9RIHsINGWgV7Ne3D8gx', '2uNpV18izVIP7DNusoWbT5', '7mXnN0kZ67wigGTn1OuNkD', '0WZ7IgzdjPvwFdjDjjuZm7', '0nDEW2hmRliyOfxJwO2CL9', '0nVQLK3LSM9SFP5xp64D7a', '0adetEU8Vg0fU96nkB9Zel', '2nrn5Hfpxg0SfuVLbnHwoL', '4x2osw7xi8s1Jy04R6rTrn', '71pUA2TXf3JHUPUgsjLtuL', '2eRG04xbRiI1AzY8uTCySk', '7hsuVq0DQonl3PnP7qkfvP', '2aB94xAF1yUDTGLTIdkw7I')

ids_remove5 = c('1c9ivsDITvYJmYqbtF02FT', '01Z8Z9K54zewyP04ZfGLSv', '2BgyJtxoixkegtJX6HqT4w', '050lnqPL6hydVpjY4MeacK', '7xCMJBL2xqnA5EsK1zeNYZ', '4GO3ufZJ7YsEnQ3PiML2Dg', '2gOghlWpnfwGGip7S1cMuq', '3amz8sGYUMDNb2wxf4PaTj', '2mknvtcck8i82nKxDPDibv', '72ktMty33X1LsIzHdRr0U5', '2EURsXo9qlt1aMWlviGCRi', '5w834ZosnqiBBV8xXCi3oD', '3UxJwjLkGvAlRi7NbYekGI', '7nwUJBm0HE4ZxD3f5cy5ok', '4pZJw2pjpyxcqdOUmYdwSY', '69g2TelswPN1IiFDKvaoSL', '6roFdX1y5BYSbp60OTJWMd', '5vngPClqofybhPERIqQMYd', '2bA6fzP0lMAQ4kz6CF61w8', '3P33qFNGBVXl86yQYWspFj', '25NQNriVT2YbSW80ILRWJa', '5e4Dhzv426EvQe3aDb64jL', '4mxWe1mtYIYfP040G38yvS', '25IG9fa7cbdmCIy3OnuH57', '0XKOBt59crntr7HQXXO8Yz', '32vWCbZh0xZ4o9gkz4PsEU', '4L1z1IcfK7lbqx8izGHaw5', '6Sb8EHpZvyjSTT5KD9DjUZ', '3XnO697XIus1M0cMuxZjos', '0jgAONnsHxrwAlhkMUVS78', '5W5bDNCqJ1jbCgTxDD0Cb3', '0dYkMe3wK29DulSa0uR8Rq', '5iB5AWIa7qreioi0AF3Bxa', '1LFKKuzn302wp15dYH28id', '0SNdq9iJyup4XY6JbNHbt6', '4jL1usPiGDK8TD46zJP4gu')

ids_remove6 = c('5wpS6LUdQYMoAreKITVtrH', '502FY8XQYRxClWgolQ3Hf2', '0m7gTbaLi5fM22MiRaRf7G', '3lQuLlW6tYv1nm3nzyJ6Ty', '3yFnqoyp1e54tZEY4ykvNB', '5sAg1HZePcFfhrs0G8A8OP', '656TGMl6uRQSQlFRVCKt7w', '3MIwMeYV8vmgQFbbRiJTp2', '3mSAqBoXQgdlpwzWsIgBzL', '6kACVPfCOnqzgfEF5ryl0x', '0c08N8UlFjDaup9TNDaWxy', '5QsUbpxSE8lCZ5ga5rnD22', '6CXezToiGS8K6jr9kr8Muv', '14h0UcPJlGYv5Qrjp7m1jQ', '2FMZn5P3WATd7Il6FgPJNu', '1RP2UpEaRzkF0Id3JigqD8', '2rTgolMbO5tyXZvD26bO8e', '59zvXMeM3koefoJWQhBki6', '69wzuykaVXlRS5KVygESvd', '7wCjDgV6nqBsHguQXPAaIM', '1dfeR4HaWDbWqFHLkxsg1d', '0MHgLfmQdutffmvWe5XBTN', '3fMbdgg4jU18AjLCKBhRSm', '602DnpaSXJB4b9DZrvxbDc', '6PpGPIrFf3LM7Q77eR4Bts', '25Kw8f1zcn9VtUk5Nl3qrp', '2p4FqHnazRucYQHyDCdBrJ', '5IUvXYeMtVmp8YJWf0eqvf',
'4lxfqrEsLX6N1N4OCSkILp', '7xkAwz0bQTGDSbkofyQt3U', '6V7pNWhlJpD0s0bMdB1PU9', '3nqA4JV7u2zG7prO5RLlAC', '5Ne8hQEmy0eZu9hlE3j58B', '2RE8NwNxsOyuNZDD0jRxHP', '0ZGAYur5k4gJprOffC6LkP')

ids_remove7 = c('1zuJe6b1roixEKMOtyrEak', '2BTZIqw0ntH9MvilQ3ewNY', '7A9yZMTrFZcgEWAX2kBfK6', '26bcq2nyj5GB7uRr558iQg', '6GnCucI2uDaYj1j39GFxkn', '4qzZ6bMrjK27yHGVjs4Pvq', '3oAbjWINfcTSAuVUlXXMty',
'3GHovBcEWpbnwCZDPF9GpM', '70ZTdbPEcEugBNay4MvxfL', '3WrFJ7ztbogyGnTHbHJFl2', '64rxQRJsLgZwHHyWKB8fiF', '2zbWfoRgxEJj2c2MyLjDkK', '5S3EbUETKlfsBRUiBKnqer', '1Z8ODXyhEBi3WynYw0Rya6')

artists_data_cleaned <- artists_data_cleaned %>%
  filter(!artist_id %in% c(ids_remove, ids_remove2, ids_remove3, ids_remove4, ids_remove5, ids_remove6, ids_remove7))

# # see the distribution of broad genres
# artists_data_cleaned %>%
#   # count per genre, sort by descending
#   count(genre, sort = TRUE) %>%
#   # add percentage column to see distribution
#   mutate(pct = round(n / sum(n) * 100, 1))

# # check what makes up the bulk of the "Other" column (used for deciding what strings to look for)
# artists_data_cleaned %>%
#   # only keep "Other" genres
#   filter(genre == "Other") %>%
#   # count per main_genre, sort by descending
#   count(main_genre, sort = TRUE) %>%
#   # see top 30 unclassified genres
#   head(30)  
```

## Data Cleaning: hits and nonhits {.smaller}

* Drop NA values and missing data
* Convert columns into correct format (numeric or char)
* Drop rows where artist not in artists.csv (can't retrieve genre)
* Drop songs outside of 2010-2019 scope
* Drop >200 artists with incorrect release dates (also drop songs with "remaster" in name)
* Drop songs by artists that didn't have a hit in 2010-2019
* Drop songs with popularity <= 1 (indicates old song uploaded late)
* Equal number of hits and nonhits (random sample)
* Normalizing `loudness`, `tempo`, and `duration_ms`
* Adding outcome variable: `hit` (boolean/logical)
* Joined hits/nonhits into `combined_music_data`
  * Size: 7,392 (3,696 hits + 3,696 nonhits) x 16

```{r numeric-columns-and-dropNA}
#| warning: false
#| echo: false
hits_data_cleaned <- hits_data %>%
  # make sure all numeric variables are numeric
  mutate(across(all_of(c("duration_ms", "key", "mode", 
                     "acousticness", "danceability", "energy", 
                     "instrumentalness", "liveness", "loudness", 
                     "speechiness", "valence", "tempo", "time_signature")), as.numeric)) %>%
  # drop rows with missing data
  drop_na(all_of(c("release_date", "duration_ms", "acousticness", "danceability", "energy", 
                 "instrumentalness", "liveness", "loudness", "speechiness", 
                 "valence", "tempo", "key", "explicit", "mode"))) %>%
  # extract first artist ID as a string
  mutate(main_artist_id = str_extract(id_artists, "(?<=\\[')[^']+"))

nonhits_data_cleaned <- nonhits_data %>%
  # make sure all numeric variables are numeric
  mutate(across(all_of(c("duration_ms", "key", "mode", 
                     "acousticness", "danceability", "energy", 
                     "instrumentalness", "liveness", "loudness", 
                     "speechiness", "valence", "tempo", "time_signature")), as.numeric)) %>%
  # drop rows with missing data
  drop_na(all_of(c("release_date", "duration_ms", "acousticness", "danceability", "energy", 
                 "instrumentalness", "liveness", "loudness", "speechiness", 
                 "valence", "tempo", "key", "explicit", "mode"))) %>%
  # extract first artist ID as a string
  mutate(main_artist_id = str_extract(id_artists, "(?<=\\[')[^']+"))

artists_data_cleaned <- artists_data_cleaned %>%
  # only keep artists in the hits dataset
  filter(artist_id %in% hits_data_cleaned$main_artist_id)
```

```{r join-genre}
#| echo: false
hits_data_cleaned <- hits_data_cleaned %>%
  left_join(
    artists_data_cleaned %>% 
      # because artist_id is a char in artists.csv
      mutate(artist_id = as.character(artist_id)) %>%
      # only want to import genre column
      select(artist_id, genre), 
    # match by artist_id
    by = c("main_artist_id" = "artist_id")
  )

nonhits_data_cleaned <- nonhits_data_cleaned %>%
  left_join(
    artists_data_cleaned %>% 
      # because artist_id is a char in artists.csv
      mutate(artist_id = as.character(artist_id)) %>% 
      # only want to import genre column
      select(artist_id, genre),
    # match by artist_id
    by = c("main_artist_id" = "artist_id")
  )

# drop songs by artists that couldn't be matched to a genre
nonhits_data_cleaned <- nonhits_data_cleaned %>%
  drop_na(genre)

# drop songs by artists that couldn't be matched to a genre
hits_data_cleaned <- hits_data_cleaned %>%
  drop_na(genre)
```

```{r 2010s-data}
#| echo: false
hits_data_2010s <- hits_data_cleaned %>%
  # keep only tracks with release date from 2010 to 2019
  filter(year(release_date) >= 2010 & year(release_date) <= 2019)

nonhits_data_2010s <- nonhits_data_cleaned %>%
  # keep only tracks with release date from 2010 to 2019
  filter(year(release_date) >= 2010 & year(release_date) <= 2019)
```

```{r drop-remasters}
#| echo: false
hits_data_2010s <- hits_data_2010s %>% 
  # drop songs that were released as remasters (indicates song came out long before)
  filter(!str_detect(tolower(song_name), "remaster"))

nonhits_data_2010s <- nonhits_data_2010s %>% 
  # drop songs that were released as remasters (indicates song came out long before)
  filter(!str_detect(tolower(song_name), "remaster"))
```

```{r remove-data-outlier}
#| echo: false
nonhits_data_2010s <- nonhits_data_2010s %>%
  # someone uploading mixes of other songs, with some extremely long mixes making the duration_ms column unhelpful for our analysis
  filter(main_artist_id != "3GghVvugpv9nXQ2YFzZNzN")

hits_data_2010s <- hits_data_2010s %>%
  # someone uploading mixes of other songs, with some extremely long mixes making the duration_ms column unhelpful for our analysis
  filter(main_artist_id != "3GghVvugpv9nXQ2YFzZNzN")
```

```{r remove-low-popularity-and-outlier-songs}
#| echo: false
hits_data_2010s <- hits_data_2010s %>%
  # we found these to be made up almost entirely of non-2010 songs
  filter(popularity > 1) %>%
  # we found this to be a few songs that were incorrectly labeled
  filter(!song_id %in% c("161KORe4R5IjtRbia9F2FK", "79Ot5mE1cG9Ge4YQKAJT1d", "2Yx4iJo4N451Z6MKFND4km", "1I56obKWXL0LsiC9Xf2f1O", "62mV6GTTEc34fREck4vguQ", "6plNVVm5iwbc3mgrAoHRun"))
```

```{r 2010s-hit-artists}
#| echo: false
nonhits_data_2010s <- nonhits_data_2010s %>%
  # only keep the artists that had a hit in the 2010s for consistency
  filter(id_artists %in% unique(hits_data_2010s$id_artists))
```

```{r take-random-sample-of-nonhits}
#| echo: false
nonhits_data_2010s_reduced <- nonhits_data_2010s %>%
  # use slice_sample so it takes a random sample of rows
  slice_sample(n = nrow(hits_data_2010s))
```

```{r combined-dataset}
#| echo: false
combined_music_data_all <- bind_rows(
  hits_data_2010s %>% mutate(hit = TRUE), # hit = TRUE for hits
  nonhits_data_2010s_reduced %>% mutate(hit = FALSE) # hit = FALSE for nonhits
  )

combined_music_data <- combined_music_data_all %>%
  # only keep the columns we care about
  select(c(release_date, duration_ms, acousticness, danceability, energy, instrumentalness, liveness, loudness, speechiness, valence, tempo, key, explicit, mode, genre, hit))

# standardize numerical columns
combined_music_data <- combined_music_data %>%
  mutate(loudness = (loudness - min(loudness)) / (max(loudness) - min(loudness))) %>%
  mutate(tempo = (tempo - min(tempo)) / (max(tempo) - min(tempo))) %>%
  mutate(duration_ms = (duration_ms - min(duration_ms)) / (max(duration_ms) - min(duration_ms)))
```

```{r combined-all-2010s}
#| echo: false
# create unbalanced dataset of all 2010 songs for EDA
combined_all_2010s <- bind_rows(
  hits_data_2010s %>% mutate(hit = TRUE), # hit = TRUE for hits
  nonhits_data_2010s %>% mutate(hit = FALSE) # hit = FALSE for nonhits
  )

# standardize numerical columns
combined_all_2010s <- combined_all_2010s %>%
  mutate(loudness = (loudness - min(loudness)) / (max(loudness) - min(loudness))) %>%
  mutate(tempo = (tempo - min(tempo)) / (max(tempo) - min(tempo))) %>%
  mutate(duration_ms = (duration_ms - min(duration_ms)) / (max(duration_ms) - min(duration_ms)))
```

# Exploratory Data Analysis

## Hit Rate by Genre (combined dataset)

Below we can see the hit rate by genre, indicating that certain genres will be more useful than others in predicting success!

```{r distribution-and-hit-rate-by-genre}
#| echo: false
# table of genres, their number of observations, hits, hit rate, artists
combined_all_2010s %>%
  group_by(genre) %>%
  summarize(
    n = n(),
    n_hits = sum(hit == TRUE, na.rm = TRUE),
    hit_rate = round((n_hits / n) * 100,2)
  ) %>%
  # sort by hit_rate
  arrange(desc(hit_rate))
```

## Summary Statistics {.smaller}

Below is a table of summary statistics of our combined dataset for our explanatory variables.

```{r explanatory-summary-table}
#| echo: false
# get mean and standard deivations of our key explanatory variables
duration_mean <- round(mean(combined_all_2010s$duration_ms, na.rm = TRUE), 3)
duration_median <- round(median(combined_all_2010s$duration_ms, na.rm = TRUE), 3)
duration_sd <- round(sd(combined_all_2010s$duration_ms, na.rm = TRUE), 3)

acousticness_mean <- round(mean(combined_all_2010s$acousticness, na.rm = TRUE), 3)
acousticness_median <- round(median(combined_all_2010s$acousticness, na.rm = TRUE), 3)
acousticness_sd <- round(sd(combined_all_2010s$acousticness, na.rm = TRUE), 3)

danceability_mean <- round(mean(combined_all_2010s$danceability, na.rm = TRUE), 3)
danceability_median <- round(median(combined_all_2010s$danceability, na.rm = TRUE), 3)
danceability_sd <- round(sd(combined_all_2010s$danceability, na.rm = TRUE), 3)

energy_mean <- round(mean(combined_all_2010s$energy, na.rm = TRUE), 3)
energy_median <- round(median(combined_all_2010s$energy, na.rm = TRUE), 3)
energy_sd <- round(sd(combined_all_2010s$energy, na.rm = TRUE), 3)

instrumentalness_mean <- round(mean(combined_all_2010s$instrumentalness, na.rm = TRUE), 3)
instrumentalness_median <- round(median(combined_all_2010s$instrumentalness, na.rm = TRUE), 3)
instrumentalness_sd <- round(sd(combined_all_2010s$instrumentalness, na.rm = TRUE), 3)

liveness_mean <- round(mean(combined_all_2010s$liveness, na.rm = TRUE), 3)
liveness_median <- round(median(combined_all_2010s$liveness, na.rm = TRUE), 3)
liveness_sd <- round(sd(combined_all_2010s$liveness, na.rm = TRUE), 3)

loudness_mean <- round(mean(combined_all_2010s$loudness, na.rm = TRUE), 3)
loudness_median <- round(median(combined_all_2010s$loudness, na.rm = TRUE), 3)
loudness_sd <- round(sd(combined_all_2010s$loudness, na.rm = TRUE), 3)

speechiness_mean <- round(mean(combined_all_2010s$speechiness, na.rm = TRUE), 3)
speechiness_median <- round(median(combined_all_2010s$speechiness, na.rm = TRUE), 3)
speechiness_sd <- round(sd(combined_all_2010s$speechiness, na.rm = TRUE), 3)

valence_mean <- round(mean(combined_all_2010s$valence, na.rm = TRUE), 3)
valence_median <- round(median(combined_all_2010s$valence, na.rm = TRUE), 3)
valence_sd <- round(sd(combined_all_2010s$valence, na.rm = TRUE), 3)

tempo_mean <- round(mean(combined_all_2010s$tempo, na.rm = TRUE), 3)
tempo_median <- round(median(combined_all_2010s$tempo, na.rm = TRUE), 3)
tempo_sd <- round(sd(combined_all_2010s$tempo, na.rm = TRUE), 3)

explicit_mean <- round(mean(combined_all_2010s$explicit, na.rm = TRUE), 3)
explicit_median <- round(median(combined_all_2010s$explicit, na.rm = TRUE), 3)
explicit_sd <- round(sd(combined_all_2010s$explicit, na.rm = TRUE), 3)

# add stats to a table
summary_stats <- tibble(
  # column of the variable names
  variable = c("duration_ms", "acousticness", "danceability", "energy",
               "instrumentalness", "liveness", "loudness",
               "speechiness", "valence", "tempo", "explicit"),
  # column of their means
  mean = c(duration_mean, acousticness_mean, danceability_mean, energy_mean,
           instrumentalness_mean, liveness_mean, loudness_mean,
           speechiness_mean, valence_mean, tempo_mean, explicit_mean),
  # column of their medians
  median = c(duration_median, acousticness_median, danceability_median, energy_median,
           instrumentalness_median, liveness_median, loudness_median,
           speechiness_median, valence_median, tempo_median, explicit_median),
  # column of their standard deviations
  standard_deviation = c(duration_sd, acousticness_sd, danceability_sd, energy_sd,
           instrumentalness_sd, liveness_sd, loudness_sd,
           speechiness_sd, valence_sd, tempo_sd, explicit_sd)
)

summary_stats
```

- Key Takeaways:
  - These statistics are post-normalization
  - Explicit is binary: ~12% of our songs are listed as explicit 
  - The distribution of duration is very tight (most songs released at this time had similar lengths)

## Trends in Energy and hit status {.smaller}

```{r mean-energy}
#| echo: false
# get mean energy for hits vs. non-hits by half-year
plot_4_data <- combined_all_2010s %>% 
  mutate(
    year_val = year(release_date),
    half = if_else(month(release_date) <= 6, "H1", "H2"),
    half_label = paste(year_val, half),
    half_index = year_val + if_else(half == "H1", 0, 0.5)
  ) %>%
  group_by(hit, year_val, half, half_label, half_index) %>%
  summarize(
    mean_energy = mean(energy, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(plot_4_data, aes(x = half_index,
                        y = mean_energy,
                        color = factor(hit),
                        group = hit)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(
    breaks = seq(2010, 2019, by = 1),
    labels = seq(2010, 2019, by = 1)
  ) +
  labs(
    title = "Average Energy for Hits vs. Non-Hits over Time",
    x = "Year",
    y = "Average Energy",
    color = "Hit"
  ) +
  theme_minimal()
```

## Trends in Valence and hit status {.smaller}

```{r mean-valence}
#| echo: false
# get the mean valence for hits vs. non-hits by year
plot_3_data <- combined_all_2010s %>% 
  mutate(
    year_val = year(release_date),
    half = if_else(month(release_date) <= 6, "H1", "H2"),
    half_label = paste(year_val, half),
    half_index = year_val + if_else(half == "H1", 0, 0.5)
  ) %>%
  group_by(hit, year_val, half, half_label, half_index) %>%
  summarize(
    mean_valence = mean(valence, na.rm = TRUE),
    .groups = "drop"
  )
# plot the trend of valence for hits vs. non-hits
ggplot(plot_3_data, aes(x = half_index,
                        y = mean_valence,
                        color = factor(hit),
                        group = hit)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(
    breaks = seq(2010, 2019, by = 1),
    labels = seq(2010, 2019, by = 1)
  ) +
  labs(
    title = "Average Valence for Hits vs. Non-Hits over Time",
    x = "Year",
    y = "Average Valence",
    color = "Hit"
  ) +
  theme_minimal()
```

## Trends in Danceability and hit status {.smaller}

```{r mean-dance}
#| echo: false
# get mean dance for hits vs. non-hits by half-year
plot_6_data <- combined_all_2010s %>% 
  mutate(
    year_val = year(release_date),
    half = if_else(month(release_date) <= 6, "H1", "H2"),
    half_label = paste(year_val, half),
    half_index = year_val + if_else(half == "H1", 0, 0.5)
  ) %>%
  group_by(hit, year_val, half, half_label, half_index) %>%
  summarize(
    mean_dance = mean(danceability, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(plot_6_data, aes(x = half_index,
                        y = mean_dance,
                        color = factor(hit),
                        group = hit)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(
    breaks = seq(2010, 2019, by = 1),
    labels = seq(2010, 2019, by = 1)
  ) +
  labs(
    title = "Average Danceability for Hits vs. Non-Hits over Time",
    x = "Year",
    y = "Average Danceability",
    color = "Hit"
  ) +
  theme_minimal()
```

## Trends in Acousticness and hit status {.smaller}

```{r mean-acousticness}
#| echo: false
# same for mean acousticness
plot_9_data <- combined_all_2010s %>% 
  mutate(
    year_val = year(release_date),
    half = if_else(month(release_date) <= 6, "H1", "H2"),
    half_label = paste(year_val, half),
    half_index = year_val + if_else(half == "H1", 0, 0.5)
  ) %>%
  group_by(hit, year_val, half, half_label, half_index) %>%
  summarize(
    mean_acousticness = mean(acousticness, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(plot_9_data, aes(x = half_index,
                        y = mean_acousticness,
                        color = factor(hit),
                        group = hit)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(
    breaks = seq(2010, 2019, by = 1),
    labels = seq(2010, 2019, by = 1)
  ) +
  labs(
    title = "Average Acousticness for Hits vs. Non-Hits over Time",
    x = "Year",
    y = "Average Acousticness",
    color = "Hit"
  ) +
  theme_minimal()
```


## Hypotheses {.smaller}

#### Hypotheses:
- Song elements that will significantly increase the model's likelihood of predicting a hit:
  - Higher `danceability`, `valence` and `loudness` 
  - Lower `duration_ms`, `acousticness`, `speechiness`, `liveness` and `instrumentalness` 
  
- Song elements that will have little to no effect:
  - `key`, `tempo`, `explicit`, and `mode` 

## Variable Correlation {.smaller}

We made the following image that demonstrates the correlation between the relevant explanatory variables in our dataset.

```{r correlation-heatmap}
#| echo: false
# select the relevant numerical columns
numerical_cols <- combined_all_2010s %>%
  select(duration_ms, acousticness, danceability, energy, 
         instrumentalness, liveness, loudness, speechiness, 
         valence, tempo)

# compute the correlation matrix
cor_matrix <- cor(numerical_cols)

# visual correlation heatmap with smaller labels
ggcorrplot(cor_matrix, 
           lab = TRUE,
           lab_size = 3)
```

# Inference and Modeling

## LASSO and Ridge Classification Models {.smaller}

- Penalizing variables (dealing with multicollinearity)
- Easy to interpret coefficients
- Prevents overfitting (higher bias, lower variance)
- k-fold cross validation: k = 5

```{python column-transf}
#| echo: false
import pandas as pd
from sklearn.compose import make_column_transformer
from sklearn.preprocessing import StandardScaler, OneHotEncoder, PolynomialFeatures

X = r.combined_music_data.drop(columns=['hit', 'release_date'])
y = r.combined_music_data['hit'].astype(int)

categorical_columns = ['key', 'explicit', 'mode', 'genre']

numerical_columns = (
  ['duration_ms', 'acousticness', 'danceability', 'energy', 'instrumentalness',\
  'liveness', 'loudness', 'speechiness', 'valence', 'tempo']
  )

preprocessor = make_column_transformer(
  (OneHotEncoder(drop="first"), categorical_columns),
  (StandardScaler(), numerical_columns),
  remainder = 'passthrough',
  verbose_feature_names_out=False, # avoid prepending preprocessor names
)

transformed_X = preprocessor.fit_transform(X)
# to make it easier to view
transformed_X = pd.DataFrame(
transformed_X,
columns = preprocessor.get_feature_names_out())
```

```{python tr-te-split}
#| echo: false
from sklearn.model_selection import train_test_split
# create split
X_train, X_test, y_train, y_test = train_test_split(
  X, y, test_size = 0.2, stratify=y, random_state = 244)
```

```{python pipe-lasso}
#| echo: false
from sklearn.linear_model import ElasticNetCV, LogisticRegressionCV
from sklearn.pipeline import Pipeline

pipeline_lasso = Pipeline(
  [
    ('preprocess', preprocessor), 
    ('estimator', LogisticRegressionCV(
    penalty="l1",
    solver="liblinear",
    cv = 5, # cross validation
    random_state=42,
    max_iter=1000))
  ]
)
```

```{python fit-lasso}
#| echo: false
#| warning: false
fitted_lasso = pipeline_lasso.fit(X_train, y_train)
```

## LASSO coefficient results {.smaller}

```{python lasso-coeff-df}
#| echo: false
lasso_coef_df = pd.DataFrame({
  "Feature": pipeline_lasso['preprocess'].get_feature_names_out(),
  "Lasso": pipeline_lasso['estimator'].coef_[0]
}).sort_values('Lasso', key=abs, ascending=False)
lasso_coef_df[:10]
```

## LASSO Accuracy results

```{python lasso-accuracy}
#| echo: false
from sklearn.metrics import accuracy_score, f1_score, roc_auc_score

# Lasso model predictions
y_pred_train_l = pipeline_lasso.predict(X_train)
y_pred_test_l = pipeline_lasso.predict(X_test)

print("Accuracy score: " +str(round(accuracy_score(y_test, y_pred_test_l)*100,2))+"%")
print("F1 Score: " +str(round(f1_score(y_test, y_pred_test_l),2)))
print("ROC-AUC Score: " +str(round(roc_auc_score(y_test, pipeline_lasso.predict_proba(X_test)[:, 1]),2)))
```

## Ridge coefficient results {.smaller}

```{python pipe-ridge}
#| echo: false
import numpy as np
from sklearn.linear_model import RidgeCV
from sklearn.linear_model import LogisticRegression

alphas_list = 10 ** np.linspace(-2, 3, 20)

# making a pipeline for ridge (logistic regression with l2 penalty)
pipeline_ridge = Pipeline(
  [
    ('preprocess', preprocessor), 
    ('estimator', LogisticRegressionCV(
      penalty='l2', 
      cv = 5, # cross validation
      max_iter=1000, 
      random_state=244))
  ]
)
```

```{python fit-ridge}
#| echo: false
#| warning: false
# fit ridge model 
pipeline_ridge.fit(X_train, y_train);
```

```{python create-coefficient-dataframe-ridge}
#| echo: false
ridge_coef_df = pd.DataFrame({
  "Feature": pipeline_ridge['preprocess'].get_feature_names_out(),
  "Ridge": pipeline_ridge['estimator'].coef_[0]
}).sort_values('Ridge', key=abs, ascending=False)
ridge_coef_df[:10]
```

## Ridge Accuracy results

```{python ridge-predictions}
#| echo: false
# Ridge model predictions
y_pred_train_r = pipeline_ridge.predict(X_train)
y_pred_test_r = pipeline_ridge.predict(X_test)

print("Accuracy score: " +str(round(accuracy_score(y_test, y_pred_test_r)*100,2))+"%")
print("F1 Score: " +str(round(f1_score(y_test, y_pred_test_r),2)))
print("ROC-AUC Score: " +str(round(roc_auc_score(y_test, pipeline_ridge.predict_proba(X_test)[:, 1]),2)))
```

```{python column-transf-no-genre}
#| echo: false
X = r.combined_music_data.drop(columns=['hit', 'release_date', 'genre'])
y = r.combined_music_data['hit'].astype(int)

categorical_columns = ['key', 'explicit', 'mode']

numerical_columns = (
  ['duration_ms', 'acousticness', 'danceability', 'energy', 'instrumentalness',\
  'liveness', 'loudness', 'speechiness', 'valence', 'tempo']
  )

preprocessor = make_column_transformer(
  (OneHotEncoder(drop="first"), categorical_columns),
  (StandardScaler(), numerical_columns),
  remainder = 'passthrough',
  verbose_feature_names_out=False, # avoid prepending preprocessor names
)

transformed_X = preprocessor.fit_transform(X)
# to make it easier to view
transformed_X = pd.DataFrame(
transformed_X,
columns = preprocessor.get_feature_names_out())
```

```{python tr-te-split-no-genre}
#| echo: false
# create split
X_train, X_test, y_train, y_test = train_test_split(
  X, y, test_size = 0.2, stratify=y, random_state = 244)
```

```{python pipe-lasso-no-genre}
#| echo: false
pipeline_lasso = Pipeline(
  [
    ('preprocess', preprocessor), 
    ('estimator', LogisticRegressionCV(
    penalty="l1",
    solver="liblinear",
    cv = 5, # cross validation
    random_state=42,
    max_iter=1000))
  ]
)
```

```{python fit-lasso-no-genre}
#| echo: false
#| warning: false
fitted_lasso = pipeline_lasso.fit(X_train, y_train)
```

## LASSO coefficient results (no genre) {.smaller}

```{python lasso-coeff-df-no-genre}
#| echo: false
lasso_coef_df = pd.DataFrame({
  "Feature": pipeline_lasso['preprocess'].get_feature_names_out(),
  "Lasso": pipeline_lasso['estimator'].coef_[0]
}).sort_values('Lasso', key=abs, ascending=False)
lasso_coef_df[:10]
```

## LASSO Accuracy results (no genre)

```{python lasso-accuracy-no-genre}
#| echo: false
# Lasso model predictions
y_pred_train_l = pipeline_lasso.predict(X_train)
y_pred_test_l = pipeline_lasso.predict(X_test)

print("Accuracy score: " +str(round(accuracy_score(y_test, y_pred_test_l)*100,2))+"%")
print("F1 Score: " +str(round(f1_score(y_test, y_pred_test_l),2)))
print("ROC-AUC Score: " +str(round(roc_auc_score(y_test, pipeline_lasso.predict_proba(X_test)[:, 1]),2)))
```

## Ridge coefficient results (no genre) {.smaller}

```{python pipe-ridge-no-genre}
#| echo: false
alphas_list = 10 ** np.linspace(-2, 3, 20)

# making a pipeline for ridge (logistic regression with l2 penalty)
pipeline_ridge = Pipeline(
  [
    ('preprocess', preprocessor), 
    ('estimator', LogisticRegressionCV(
      penalty='l2', 
      cv = 5, # cross validation
      max_iter=1000, 
      random_state=244)
      )
  ]
)
```

```{python fit-ridge-no-genre}
#| echo: false
#| warning: false
# fit ridge model 
pipeline_ridge.fit(X_train, y_train);
```

```{python create-coefficient-dataframe-ridge-no-genre}
#| echo: false
ridge_coef_df = pd.DataFrame({
  "Feature": pipeline_ridge['preprocess'].get_feature_names_out(),
  "Ridge": pipeline_ridge['estimator'].coef_[0]
}).sort_values('Ridge', key=abs, ascending=False)
ridge_coef_df[:10]
```

## Ridge Accuracy results (no genre)

```{python ridge-predictions-no-genre}
#| echo: false
# Ridge model predictions
y_pred_train_r = pipeline_ridge.predict(X_train)
y_pred_test_r = pipeline_ridge.predict(X_test)

print("Accuracy score: " +str(round(accuracy_score(y_test, y_pred_test_r)*100,2))+"%")
print("F1 Score: " +str(round(f1_score(y_test, y_pred_test_r),2)))
print("ROC-AUC Score: " +str(round(roc_auc_score(y_test, pipeline_ridge.predict_proba(X_test)[:, 1]),2)))
```

# Discussion and Future Work

## Conclusions + Limitations {.smaller}

- Model does not *fully* capture what makes a song a hit, but it does a pretty decent job
  - Achieved around 70% accuracy
  
- Genre is highly influential
  - Model performance increases by ~3% when genre is included
  - Spotify-derived characteristics are not as important

- Dates may not reflect reality
  - Unlabeled remasters released during the sample period may bias results
    - These songs may not follow the 2010s "formula for success"
  - Certain songs have incorrect dates due to when they were added to music streaming platforms
  - Potential use of Spotify API to get correct dates

- Varying amount of data by year/genre
  - Different numbers of observations per year/genre may also be skewing data

## Future Work {.smaller}
- Look further into genres
  - Certain genres with low numbers of observations may be biasing our current estimates

- Test model on currently successful songs/new releases
  - See how accurate the model is with music from outside this period
  
- Potentially looking at other predictors
  - Collaborations vs. Solo Song
  - Whether the artist is signed vs. independent
    - Could help proxy some of the "marketing" impact
  - A song's language
  
# Thank you :)

